A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2012-03-06 papers/contact/paper.tex
Fixed bugs to WBm2 derivs in paper and code Tested everything on crystal.
clean up bits of contact paper Tested everything on bingley.

We will define our terms by making use of the two-particle density
$n^{(2)}(\rr_1,\rr_2)$, which gives the probability per unit volume of
finding one particle at position $\rr_1$ and the other at position
$\rr_2$.  The pair correlation function is defined by
\begin{align}
  g(\rr_1,\rr_2) &\equiv \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
\end{align}
In a homogeneous fluid, the pair correlation only depends on the
distance $|\rr_1-\rr_2|$, and can be expressed as a function of a
single variable, with the contact value being when that distance is
$2R$.  It is desirable for reasons of efficiency to limit CDFT
functionals to one-center convolutions, which leads us to seek a
simplified expression for the contact value of the correlation
function---which is the same as the contact value of the cavity
correlation function for hard spheres.
%
\fixme{The following bit is awkward.}
%
In a system with an inhomogeneous density, we seek a \emph{local}
contact value of the correlation function.  There are two reasonable
options for defining such a local function:  a symmetric (which we refer to as S) and an
asymmetric (which we refer to as A) formulation.

For the S case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{n^S_\textit{contact}(\rr)}{n_S(\rr)}
\end{align}
\begin{align}\label{eq:n-contactS}
  n^S_\textit{contact}(\rr) &= \frac1{4\pi R^2}\int n^{(2)}(\rr - \rr', \rr + \rr')\delta(|\rr'| -
  R)d\rr'
\end{align}
\begin{align}
  n_S(\rr) &= \frac{1}{4\pi R^2}\int n(\rr')\delta(|\rr-\rr'|-R) d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_S$ is the same as the density $n_0$
defined in Fundamental Measure Theory (FMT), and seems ideal for
treating touching spheres.

And for the A case:
\begin{align}
  g^A_\sigma(\rr) &= \frac{n^A_\textit{contact}(\rr)}{n_A(\rr)} \\
  n^A_\textit{contact}(\rr) &= \frac1{16\pi R^2}\int n^{(2)}(\rr, \rr + \rr')
  \delta(|\rr'| - 2R)d\rr' \\
  n_A(\rr) &= \frac1{16\pi R^2}\int n(\rr') \delta(|\rr-\rr'|-2R) d\rr'
\end{align}
In this case, we find the average number of contacts for a sphere at a
given position.

In the process of defining these two averaged correlation functions,
we also defined four averaged densities, two \emph{contact
  densities}, $n^S_\textit{contact}(\rr)$ and $n^A_\textit{contact}(\rr)$,
and two \emph{contacting densities}, $n_S(\rr)$ and $n_A(\rr)$.  The
former are the density of spheres that are in contact with each other,
while the latter are the density of spheres available to be in
contact.  In the homogeneous limit, $n_S = n_A = n$ and
$n^S_\textit{contact} = n^A_\textit{contact} = g(\sigma)n$.  It is an
open question, which of these averages will be more useful in any
particular functional.  Suffice to say, either average is a
\emph{possible} way to convert a function that is defined for a
homogeneous system to a functional that is applicable to inhomogeneous
systems.

@@

\begin{align} 
A_{HS} &= \int \left\{
    -n_0 \ln\left(1 - n_3\right)
    + (1 + \frac{1}{9}n_3^2 \phi_2(n_3))   \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
    + (1 - \frac{4}{9}n_3\phi_3(n_3))\frac{(n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})}{24\pi (1-n_3)^2}
    \right\}
\end{align}
 where
 \begin{align}
   \phi_2 &= (6n_3 - 3n_3^2 + 6(1-n_3)\ln(1-n_3))\frac{1}{n_3^3}\\
   \phi_3 &= (6n_3 - 9n_3^2 + 6n_3^3 + 6(1-n_3)^2 \ln(1-n_3))\frac{1}{4n_3^3}
 \end{align}
For the following derivitives we have
\begin{align}
   \frac{\delta \phi_2}{\delta n_3} &= ((-6+4n_3)\ln(1-n_3) - 6n_3 + n_3^2]\frac{3}{n_3^4}\\
   \frac{\delta \phi_3}{\delta n_3} &= (-6n_3 + 5n_3^2 + (-2n_3^2 + 8n_3 -6)\ln(1-n_3))\frac{3}{4n_3^4}
 \end{align}
And using these
\begin{align}
  \frac{\delta A_{HS}}{\delta n_3} &= \psi _1 + \psi _2 + \psi _3
\end{align}
where
\begin{align}
  \psi_1 &= \frac{n_0}{1-n_3}\\
  \psi_2 &= (\frac{n_1n_2 - \mathbf{n}_1\cdot \mathbf{n}_2}{1-n_3})(\frac{2}{9}n_3\phi_2 + \frac{1}{9}n_3^2\frac{\delta \phi_2}{\delta n_3}) + (1 + \frac{1}{9}n_3^2\phi_2)(\frac{n_1n_2 - \mathbf{n}_1\cdot \mathbf{n}_2}{(1-n_3)^2})\\
  \psi_3 &= (\frac{n_2^3 - 3n_2\mathbf{n}_2 \cdot \mathbf{n}_2}{24\pi(1-n_3)^2})(-\frac{4}{9}\phi_3 - \frac{4}{9}n_3\frac{\delta \phi_3}{\delta n_3}) + 2(1 - \frac{4}{9}n_3\phi_3)(\frac{n_2^3 - 3n_2\mathbf{n}_2 \cdot \mathbf{n}_2}{24\pi(1-n_3)^3}) 
\end{align}
We also have
\begin{align}
  \frac{\delta A_{HS}}{\delta n_2} &= (1 + \frac{1}{9}n_3^2\phi_2(n_3))\frac{n_1}{1-n_3} + (1 - \frac{4}{9}n_3\phi_3)(\frac{3n_2^2 - 3\mathbf{n}_2 \cdot \mathbf{n}_2}{24\pi(1 - n_3)^2})\\
  \frac{\delta A_{HS}}{\delta n_1} &= (1 + \frac{1}{9}n_3^2\phi_2)\frac{n_2}{1 - n_3}\\
  \frac{\delta A_{HS}}{\delta n_0} &= -\ln(1 - n_3)\\
  \frac{\delta A_{HS}}{\delta \mathbf{n}_1} &= -(1 + \frac{1}{9}n_3^2 \phi_2)\frac{\mathbf{n}_2}{1 - n_3}\\
  \frac{\delta A_{HS}}{\delta \mathbf{n}_2} &= -(1 + \frac{1}{9}n_3^2\phi_2)\frac{\mathbf{n}_1}{1 - n_3} 
  - (1 - \frac{4}{9}n_3\phi_3)\frac{6\mathbf{n}_2}{24\pi(1 - n_3)^2}
\end{align}

@@

\section{Gross' functional}\label{sec:gross}
We compare with the functional of Gross \emph{et
  al}\cite{gross2009density}, which is of the second kind ($g^b$), where the
contact density is evaluated at the sphere location.
\begin{align}
  n_\textit{contact} &= n_b \frac{1 - \frac12 n_{3'}}{\left(1 -
    n_{3'}\right)^3} \\
  n_{3'}(\rr) &= \int n(\rr')\Theta(|\rr-\rr'| - 2R) d\rr'
\end{align}
where $n_b$ is defined in the introduction and $n_{3'}$ is an averaged
filling fraction defined in a manner analogous to the $n_3$ of FMT,
but with the hard-sphere radius replaced by the hard-sphere diameter,
so the integrals range over spheres that are twice as large.

@@

>>>>

We will define our terms by making use of the two-particle density
$n^{(2)}(\rr_1,\rr_2)$, which gives the probability per unit volume of
finding one particle at position $\rr_1$ and the other at position
$\rr_2$.  The pair correlation function is defined by
\begin{align}
  g(\rr_1,\rr_2) &\equiv \frac{n^{(2)}(\rr_1,\rr_2)}{n(\rr_1)n(\rr_2)}
\end{align}
In a homogeneous fluid, the pair correlation only depends on the
distance $|\rr_1-\rr_2|$, and can be expressed as a function of a
single variable, with the contact value being when that distance is
the diameter $\sigma$.  It is desirable for reasons of efficiency to limit CDFT
functionals to one-center convolutions, which leads us to seek a
simplified expression for the contact value of the correlation
function---which is the same as the contact value of the cavity
correlation function for hard spheres.
In a system with an inhomogeneous density, we seek a \emph{local}
value for $g_\sigma$.  There are two reasonable options for defining
such a local function: a symmetric formulation (which we refer to as $S$) and an
asymmetric formulation (which we refer to as $A$). \fixme{Should we
  introduce a third one, for Yu and Wu?}

For the symmetric $S$ case:
\begin{align}
  g^S_\sigma(\rr) &= \frac{1}{n_S(\rr)^2}\int n^{(2)}(\rr - \rr', \rr
  + \rr')
  \frac{\delta(|\rr'| -\sigma/2)}{\pi\sigma^2}d\rr' \\
  n_S(\rr) &= \int n(\rr')\frac{\delta(|\rr-\rr'|-\sigma/2)}{\pi\sigma^2} d\rr'
\end{align}
This functional gives a value averaged over spheres that touch at the
position $\rr$.  The density $n_S$ is the same as the density $n_0$
defined in Fundamental Measure Theory (FMT), and seems ideal for
treating touching spheres.

And for the asymmetric $A$ case:
\begin{align}
  g^A_\sigma(\rr) &= \frac{1}{n(\rr)n_A(\rr)}
  \int n^{(2)}(\rr, \rr + \rr')
  \frac{\delta(|\rr'| - \sigma)}{4\pi\sigma^2}d\rr' \\
  n_A(\rr) &= \int n(\rr')
  \frac{\delta(|\rr-\rr'|-\sigma)}{4\pi\sigma^2} d\rr'
\end{align}
In this case, the two-particle density is averaged over spheres
touching a sphere that is aat the given position $\rr$.

In the process of defining these two averaged correlation functions,
we also defined two averaged densities, $n_S(\rr)$ and $n_A(\rr)$,
which are the density of spheres available to be in
contact.  In the homogeneous limit, $n_S = n_A = n$.  It is an
open question, which of these averages will be more useful in any
particular functional.  Suffice to say, either average is a
\emph{possible} way to convert a function that is defined for a
homogeneous system to a functional that is applicable to inhomogeneous
systems.

@@

\begin{align} 
A_{HS} &= \int \left\{
    -n_0 \ln\left(1 - n_3\right)
    + \left(1 + \frac{1}{9}n_3^2 \phi_2(n_3)\right)   \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
    + (1 - \frac{4}{9}n_3\phi_3(n_3))\frac{(n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})}{24\pi (1-n_3)^2}
    \right\}
\end{align}
 where
 \begin{align}
   \phi_2 &= (6n_3 - 3n_3^2 + 6(1-n_3)\ln(1-n_3))\frac{1}{n_3^3}\\
   \phi_3 &= (6n_3 - 9n_3^2 + 6n_3^3 + 6(1-n_3)^2 \ln(1-n_3))\frac{1}{4n_3^3}
 \end{align}
For the following derivitives we have
\begin{align}
   \frac{\delta \phi_2}{\delta n_3} &= ((-6+4n_3)\ln(1-n_3) - 6n_3 + n_3^2]\frac{3}{n_3^4}\\
   \frac{\delta \phi_3}{\delta n_3} &= (-6n_3 + 5n_3^2 + (-2n_3^2 + 8n_3 -6)\ln(1-n_3))\frac{3}{4n_3^4}
 \end{align}
And using these
\begin{align}
  \frac{\delta A_{HS}}{\delta n_3} &= \psi _1 + \psi _2 + \psi _3
\end{align}
where
\begin{align}
  \psi_1 &= \frac{n_0}{1-n_3}\\
  \psi_2 &= (\frac{n_1n_2 - \mathbf{n}_1\cdot \mathbf{n}_2}{1-n_3})(\frac{2}{9}n_3\phi_2 + \frac{1}{9}n_3^2\frac{\delta \phi_2}{\delta n_3}) + (1 + \frac{1}{9}n_3^2\phi_2)(\frac{n_1n_2 - \mathbf{n}_1\cdot \mathbf{n}_2}{(1-n_3)^2})\\
  \psi_3 &= (\frac{n_2^3 - 3n_2\mathbf{n}_2 \cdot \mathbf{n}_2}{24\pi(1-n_3)^2})(-\frac{4}{9}\phi_3 - \frac{4}{9}n_3\frac{\delta \phi_3}{\delta n_3}) + 2(1 - \frac{4}{9}n_3\phi_3)(\frac{n_2^3 - 3n_2\mathbf{n}_2 \cdot \mathbf{n}_2}{24\pi(1-n_3)^3}) 
\end{align}
We also have
\begin{align}
  \frac{\delta A_{HS}}{\delta n_2} &= (1 + \frac{1}{9}n_3^2\phi_2)\frac{n_1}{1-n_3} + (1 - \frac{4}{9}n_3\phi_3)(\frac{3n_2^2 - 3\mathbf{n}_2 \cdot \mathbf{n}_2}{24\pi(1 - n_3)^2})\\
  \frac{\delta A_{HS}}{\delta n_1} &= (1 + \frac{1}{9}n_3^2\phi_2)\frac{n_2}{1 - n_3}\\
  \frac{\delta A_{HS}}{\delta n_0} &= -\ln(1 - n_3)\\
  \frac{\delta A_{HS}}{\delta \mathbf{n}_1} &= -(1 + \frac{1}{9}n_3^2 \phi_2)\frac{\mathbf{n}_2}{1 - n_3}\\
  \frac{\delta A_{HS}}{\delta \mathbf{n}_2} &= -(1 + \frac{1}{9}n_3^2\phi_2)\frac{\mathbf{n}_1}{1 - n_3} 
  - (1 - \frac{4}{9}n_3\phi_3)\frac{6n_2\mathbf{n}_2}{24\pi(1 - n_3)^2}
\end{align}

@@

\section{Gross' functional}\label{sec:gross}
We compare with the functional of Gross \emph{et
  al}\cite{gross2009density}, which is of the second kind ($g^b$), where the
contact density is evaluated at the sphere location.
\begin{align}
  n_\textit{contact} &= n_{A} \frac{1 - \frac12 n_{3'}}{\left(1 -
    n_{3'}\right)^3} \\
  n_{3'}(\rr) &= \int n(\rr')\Theta(|\rr-\rr'| - 2R) d\rr'
\end{align}
where $n_A$ is defined in the introduction and $n_{3'}$ is an averaged
filling fraction defined in a manner analogous to the $n_3$ of FMT,
but with the hard-sphere radius replaced by the hard-sphere diameter,
so the integrals range over spheres that are twice as large.

@@

The functional of Yu and Wu for $g_\sigma$ is a symmetric formulation,
but instead of using $n_0$ as the associated density, it uses a
density given by
\begin{align}
  n_{YW} &= n_0 \zeta \\
  \zeta &= 1 - \frac{\mathbf{n_2}\cdot\mathbf{n_2}}{n_2^2}
\end{align}
where the function $\zeta$ is a measure of local inhomogeneity at the
point of contact.  Given this definition of density, $g_\sigma$ is
given by:
\begin{align}
  g_\sigma^{YW} &= \frac{1}{1-n_3}
    + \frac32 \frac{R n_2}{3}\frac{\zeta}{(1-n_3)^2}
    + \frac12 \frac{\left(\frac{R n_2}{3}\right)^2 \zeta}{(1-n_3)^3}
    \\
  g_\sigma^{YW} &= \frac{1}{1-n_3}
    + \frac12 \frac{R n_2\zeta}{(1-n_3)^2}
    + \frac1{18} \frac{R^2 n_2^2 \zeta}{(1-n_3)^3}
\end{align}
We should write a bit more about this functional.  I think we can view
it as a third alternative, which we could also predict:
\begin{align}
  g_\sigma^{S'}(\rr) &= \frac1{n_{S'}(\rr)^2}\frac1{4\pi R^2}\int n^{(2)}(\rr - \rr', \rr + \rr')\delta(|\rr'| -
  R)d\rr'
\end{align}
\begin{align}
  n_{S'}(\rr) &= \zeta(\rr) \frac{1}{4\pi R^2}\int n(\rr')\delta(|\rr-\rr'|-R) d\rr'
\end{align}
This is closely related to $g_\sigma^{S}$, but is ever so slightly
different.  In order to be completely fair, we should see about
computing this from our MC calculations.  Unfortunately, it will
require that we compute $\mathbf{n}_2 \cdot \mathbf{n}_2$ which will
be a bit of a pain.  $\ddot\frown$
